---
title: "PL Injuries Data Prep"
author: "Tamas Koncz"
date: "July 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)

library(caret)

rm(list = ls())
injuries_raw <- fread("./../Data/injury_soccer/injury_data_pg.csv") 
```


Import UDFs:
```{r}
#EDA
source("f_sec_scale.R")
source("f_breakdown_by_bins.R")
source("f_breakdown_by_variable.R")

#Modeling
source("fun_calc_auc.R")
source("fun_roc_create.R")
source("fun_roc_create.R")
source("fun_calibration_create.R")
```


```{r}
injuries <- injuries_raw %>%
                  select(Date,                 `Game week`,      `Kick-off`,        mid,
                         pid,                  `First name`,     `Last name`,       minutes,
                         injured,               injury_length,    injury_type,
                         `Country of birth`,    Nationality,     `Date of birth`,
                         Position,              Foot,             Height,           Weight,
                         starts_with("all_"),   starts_with("pl_"))
rm(injuries_raw)
#injuries %>% glimpse()
```


Formatting clean-up for date variables
```{r}
injuries <- injuries %>%
              mutate(Date = as.Date(Date, "%Y-%m-%d"),
                     `Date of birth` = as.Date(`Date of birth`, "%Y-%m-%d")) %>%
              mutate(age_in_years = time_length(difftime(Date, `Date of birth`), "years")) 
```

As shown, the injury data is only complete for 2010-2017. That should be still plenty for statistical analysis.
```{r}
injuries %>%
  mutate(year = year(Date)) %>%
  group_by(year) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "year")
```

```{r}
injuries <- injuries %>%
              filter(year(Date) %in% c(2010:2017))
```


And for height & weight
```{r}
injuries <- injuries %>%
              filter(!(is.null(Height) | Height == "")) %>%
              mutate(`Height (cm)` = gsub(pattern = " cm", replacement = "", x = Height)) %>%
              mutate(`Height (cm)` = as.numeric(`Height (cm)`)) %>%
              select(everything(), -Height)

injuries <- injuries %>%
              filter(!(is.null(Weight) | Weight == "")) %>%
              mutate(`Weight (kg)` = gsub(pattern = " kg", replacement = "", x = Weight)) %>%
              mutate(`Weight (kg)` = as.numeric(`Weight (kg)`)) %>%
              select(everything(), -Weight)
```

Add BMI column:
```{r}
injuries <- injuries %>%
              mutate(bmi = `Weight (kg)` / (`Height (cm)` / 100)^2)
```

##Explain choice to investigate hamstring issues

By occurence:
```{r, fig.width = 8, fig.height = 5, fig.align = 'center'}
injuries %>%
  filter(injury_type != '0') %>%
  count(injury_type, sort = T) %>%
  mutate(injury_type = ifelse(n < 25, 'Other', injury_type)) %>%
  mutate(injury_type = reorder(injury_type, n)) %>%
  group_by(injury_type) %>%
  summarize(n = sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = injury_type, y = n)) +
    geom_bar(stat = "identity", fill = "lightblue") +
    geom_text(aes(label = n), hjust = 1.05, vjust = 0.325, fontface = "bold") +
    coord_flip() + 
    labs(title    = "The most common injury is to the hamstring",
         subtitle = "Number of injuries by type",
         x        = "Injury Type",
         y        = "Count of injuries",
         caption  = "Note: Injuries that happened less than 25x are group as 'Other'") + 
    theme_minimal() +
    theme(panel.grid.minor = element_blank()) +
    theme(plot.subtitle = element_text(color = "gray60"),
          plot.caption  = element_text(color = "gray60")) +
    theme(axis.title.x = element_text(hjust = 0.36))
```

By length ("severity")
```{r, fig.width = 8, fig.height = 5, fig.align = 'center'}
injuries %>%
  filter(injury_type != '0') %>%
  group_by(injury_type) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(injury_type = ifelse(n < 25, 'Other', injury_type)) %>%
  mutate(injury_type = reorder(injury_type, n)) %>%
  mutate(injury_length_capped_180 = ifelse(injury_length > 180, 180, injury_length)) %>%
  ggplot(aes(x = injury_type, y = injury_length_capped_180)) + 
    geom_boxplot() +
    coord_flip() +
    labs(title    = "There is significant variety among seriousness of injuries",
         subtitle = "Length of injuries by type",
         x        = "Injury Type",
         y        = "Length of injuries",
         caption  = "Note: Injuries that happened less than 25x are group as 'Other'") + 
    theme_minimal() +
    theme(panel.grid.minor = element_blank()) +
    theme(plot.subtitle = element_text(color = "gray60"),
          plot.caption  = element_text(color = "gray60")) +
    theme(axis.title.x = element_text(hjust = 0.36))
```



For now, focus on 'Hamstring', then see if it can be extended to all strains.
(Note: at the time of writing this - https://www.bbc.com/sport/football/44671242)
```{r}
injuries %>%
  filter(injury_type == "Hamstring") %>%
  filter(injury_length > 1) %>% #should miss at least 1 day
  mutate(injury_length_capped_180 = ifelse(injury_length >= 180, 180, injury_length)) %>% 
  ggplot(aes(x = log(injury_length_capped_180))) +
    geom_density(fill = "lightblue") +
    theme_minimal()
```

```{r}
hamstring_only <- injuries %>% 
                    filter(injury_type == "0" | injury_type == "Hamstring")

#keep only injuries with "significant" missed time
hamstring_only <- hamstring_only %>%
                    filter(!(injury_type == "Hamstring" & injury_length < 7))

rm(injuries)
```


#### Minutes played in a given game - less time, less chance to get injured?
```{r}
hamstring_only %>%
  mutate(`Minutes played` = cut(minutes, 
                                 breaks = c(0, 15, 30, 45, 60, 75,  Inf), 
                                 include.lowest = T,
                                 labels = c("-15", "15-30", "30-45", "45-60", "60-75", "75+"))) %>%
  group_by(`Minutes played`) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "`Minutes played`")
```

#### Age, weight, height, bmi  

Age: slight increase over the years for significantly large groups
```{r}
hamstring_only %>%
  mutate(`Age (Years)` = cut(age_in_years, 
                                 breaks = c(-Inf, 20, 22.5, 25, 27.5, 30, 32.5, 35, Inf), 
                                 include.lowest = T)) %>%
  group_by(`Age (Years)`) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "`Age (Years)`")
```

Weight. TODO: Better categories needed
```{r}
hamstring_only %>%
  mutate(`Weight (kg)` = cut(`Weight (kg)`, 
                              breaks = seq(0, 125, by = 5), 
                              include.lowest = T)) %>%
  group_by(`Weight (kg)`) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "`Weight (kg)`")
```

Height:
```{r}
hamstring_only %>%
  mutate(`Height (cm)` = cut(`Height (cm)`, 
                              breaks = seq(0, 220, by = 5), 
                              include.lowest = T)) %>%
  group_by(`Height (cm)`) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "`Height (cm)`")
```


BMI - does not seem to have any trend.
```{r}
hamstring_only %>%
  mutate(BMI = cut(bmi, breaks = c(-Inf, 19, 21, 23, 25, 27, Inf), include.lowest = T)) %>%
  group_by(BMI) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "BMI")
```

#### By Kick-off time  
  
Maybe a slight impact... 
```{r}
hamstring_only %>%
  mutate(kick_off_hour = as.numeric(substr(`Kick-off`, 1, 2)) + as.numeric(as.numeric(substr(`Kick-off`, 4, 5)) > 30)) %>%
  mutate(`Kick-off Hour` = cut(kick_off_hour, breaks = c(0, 15, 17, 19, 24), 
                                 include.lowest = T, labels = c("-15", "15-17", "17-19", "19-"))) %>%
  group_by(`Kick-off Hour`) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "`Kick-off Hour`")
```

#### by Year & Month  

 
```{r}
hamstring_only %>%
  mutate(Year = year(Date)) %>%
  group_by(Year) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "Year")
```

#check against weather? or other variables
```{r}
Month_mapping <- data.frame(Month_Num = c(1:12), Month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                                           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
                            Month_in_Season = c(6, 7, 8, 9, 10, 11, 12, 1, 2, 3, 4, 5))

hamstring_only %>%
  mutate(Month_Num = month(Date)) %>%
  left_join(Month_mapping, by = "Month_Num") %>%
  mutate(Month = reorder(Month, Month_in_Season)) %>%
  group_by(Month) %>%
  summarize(game_count = n(),
  injury_count = (sum(injury_type != 0))) %>%
  ungroup() %>%
  mutate(injury_rate_pct = injury_count / game_count * 100) %>%
  f_breakdown_by_bins(s = "Month")

rm(Month_mapping)
```
  
#### Comparison with categorical variables

```{r}
f_breakdown_by_variable(hamstring_only, "Position")
```

Note: Position is missing for a few records. Zero injuries. TODO: Investigate structure of missing data!
Goalkeepers getting injured a lot less.

```{r}
f_breakdown_by_variable(hamstring_only, "Foot")
```

Note: injury count is significantly lower when Foot is missing

TODO: create some groups based on country of birth / nationality ?
```{r}
f_breakdown_by_variable(hamstring_only, "Nationality")
```



#### Fixing NAs & encode categicals

NAs:
```{r}
hamstring_only <- hamstring_only %>%
                    select(mid, pid, injured, Position, Foot, age_in_years, `Height (cm)`, `Weight (kg)`, bmi,
                                     Date, starts_with("all_"), starts_with("pl_"))

hamstring_only <- hamstring_only %>% mutate(Position = ifelse(Position == "", "Missing", Position))
hamstring_only <- hamstring_only %>% mutate(Foot     = ifelse(Foot     == "", "Missing", Foot))
```

Encode categoricals:
```{r}
position_encoded <- hamstring_only %>%
                      select(mid, pid, Position) %>%
                      mutate(Position = paste("Position", Position, sep = "_")) %>%
                      mutate(i = 1) %>% tidyr::spread(key = Position, value = i, fill = 0) %>%
                      select(-Position_Goalkeeper) # assigning Goalkeeper as the "base value" for position

foot_encoded     <- hamstring_only %>%
                      select(mid, pid, Foot) %>%
                      mutate(Foot = paste("Foot", Foot, sep = "_")) %>%
                      mutate(i = 1) %>% tidyr::spread(key = Foot, value = i, fill = 0) %>%
                      select(-Foot_Right) # assigning Right as the "base value" for Foot

hamstring_only <- hamstring_only %>%
                    select(-Position, -Foot) %>%
                    left_join(position_encoded, by = c("mid", "pid")) %>%
                    left_join(foot_encoded,     by = c("mid", "pid"))

rm(position_encoded)
rm(foot_encoded)
```


Games/minutes fields are populated with NAs for 25 rows:
```{r}
#why is this data missing?
hamstring_only <- hamstring_only %>%
                    filter(!is.na(all_games_season) & 
                           !is.na(pl_games_season))
```


### Prediction  
For now, let's build two models: glm & rf. let's see what signals can we pick up  

Remove ids:
```{r}
unique_ids <- hamstring_only %>%
                select(mid, pid) %>%
                mutate(unique_id = paste(mid, pid, sep = "_"))

hamstring_only <- hamstring_only %>%
                    mutate(unique_id = paste(mid, pid, sep = "_")) %>%
                    select(-mid, -pid)
```


Caret needs the target to be a factor:
```{r}
hamstring_only$injured <- factor(hamstring_only$injured, levels = c("1", "0"), labels = c("Yes", "No"))

hamstring_only %>%
  count(injured)
```

Set aside last full years of data to evaluate performance of final model:
```{r}
performance_eval <- hamstring_only %>% 
                      filter(year(Date) == 2017) %>%
                      select(-Date)

hamstring_only <- hamstring_only %>% 
                      filter(year(Date) < 2017) %>%
                      select(-Date)
```


```{r}
training_ratio <- 0.20

set.seed(93) #for reproducibility
train_indices <- createDataPartition(y = hamstring_only[["injured"]],
                                     times = 1,
                                     p = training_ratio,
                                     list = FALSE)

data_train <- hamstring_only[train_indices, ]
data_test  <- hamstring_only[-train_indices, ]
```

Setting up cross-validation:
```{r}
train_control <- trainControl(method = "cv",
                              number = 10,
                              classProbs = TRUE,
                              summaryFunction = twoClassSummary)
```

Simple logit:
```{r}
set.seed(93)
glm_model <- train(injured   ~ .,
                   method    = "glm",
                   family    = "binomial",
                   metric    = "ROC",
                   data      = data_train %>% select(-unique_id),
                   trControl = train_control)

summary(glm_model)
```

```{r}
glm_prediction_probs <- predict.train(glm_model, newdata = data_test, type = "prob")
test_truth <- data_test$injured
```

```{r}
glm_AUC <- fun_calc_auc(glm_prediction_probs, test_truth)
glm_ROC <- fun_roc_create(glm_prediction_probs, test_truth, glm_AUC, "glm")
glm_calibration <- fun_calibration_create(glm_prediction_probs, test_truth, "glm")
```


Random Forest - this will be our benchmark model:
```{r}
# train_Control <- trainControl(method = "CV",
#                               number = 10,
#                               classProbs = TRUE,
#                               summaryFunction = twoClassSummary,
#                               selectionFunction = "best")
# 
# tune_Grid_rf <- data.frame(mtry = c(4:round(sqrt(ncol(data_train)-1))))
# 
# gc()
# memory.limit(size = 100000)
# set.seed(93)
# rf_500_fit <- train(injured ~ .,
#                    data = data_train,
#                    method = "rf",
#                    metric = "ROC",
#                    trControl = train_Control,
#                    tuneGrid = tune_Grid_rf ,
#                    ntree = 50,
#                    importance = T)
# 
# print(rf_500_fit)
# 
# saveRDS(model, "rf_500_fit.rds")
# rm(rf_500_fit)
# #my_model <- readRDS("model.rds")
```

```{r}
# rf_500_prediction_probs <- predict.train(rf_500_fit, newdata = data_test, type = "prob")
```

```{r}
# rf_500_AUC         <- fun_calc_auc(          rf_500_prediction_probs, test_truth)
# rf_500_ROC         <- fun_roc_create(        rf_500_prediction_probs, test_truth, rf_500_AUC, "glm")
# rf_500_calibration <- fun_calibration_create(rf_500_prediction_probs, test_truth, "glm")
```

```{r}
# resamples_object <- resamples(list("glm_model"  = glm_model,
#                                    "rf_500_fit" = rf_500_fit))
# 
# resamples_object$values %>%
#   tidyr::gather(key= "Resample", factor_key = F) %>%
#   setnames(c("Fold", "Model~Metric", "Value")) %>%
#   mutate(model  = stringr::str_split(`Model~Metric`,  "~", simplify = T)[, 1],
#          metric = stringr::str_split(`Model~Metric`,  "~", simplify = T)[, 2]) %>%
#   mutate(model = factor(model, levels = c("glm_model", "rf_500_fit"))) %>%
#   ggplot(aes(x= model, y= Value, fill = model)) +
#     geom_boxplot() +
#     facet_grid(~metric)
```

